Developmental plasticity to high light intensity amplifies the advantage of amphistomous leaves

### 2025-07-20 ms/manuscraps.qmd
example-aa figure

>>>>

##### Effects of light intensity, light treatment, and population on \aax

We tested for effects of light intensity, light treatment, population, and their interactions on \aax. All models included effects of light intensity and light treatment, as well as random effects of population and replicate within population. More complex models included interactions between light intensity and light treatment, as well as random effects of population on the effects of light intensity and light treatment. We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects of light intensity and treatment. We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within population. We accounted for the phylogenetic structure among the random effects of population using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997]. The expected covariance between populations $i$ and $j$ ($\text{Cov}(i, j)$ ) is:

$$\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})$$
where $\sigma^2$ is the variance of the random effect, $\alpha$ is the rate of decay of the covariance with phylogenetic distance, $D_{ij}$, the phylogenetic distance between populations $i$ and $j$. We estimated $\sigma^2 / (2 \alpha)$ and $\alpha$ as a separate parameters and reparameterized them as $\sigma^2$ and $\alpha$. We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.

We modeled the residual standard deviation of \aax, (i.e. phylogenetically unstructured variation unaccounted for by explanatory variables) on a log-link scale with effects of light intensity and light treatment. We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on the residual standard deviation intercept and a weakly informative normal prior with mean $0$ and standard deviation $1$ on the effects of light intensity and light treatment on the residual standard deviation.

##### Effects of native light habitat on population-level \aax

We tested a linear effect of native light habitat on population-level \aax. In Models 1 and 2, we used the random intercept of \aax{} as a response variable; in model 3 we used the random intercept plus the random effect of population at high light intensity; in model 4 we used the random intercept plus the random effect of population in the sun treatment; in model 5 we used the random intercept plus the random effects of population at high light intensity and sun treatment. We used a weakly informative normal prior with mean $0$ and standard deviation $1$ for the slope and intercept. We accounted for the phylogenetic structure among the model residuals using an (OU) process. We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.

## note on LOOIC

We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.

### 2025-07-20 ms/ms.qmd
example-aa figure

#### \agcurve{} curve parameters

@@

#### \aax{} for each light intensity within leaf using \agcurve{} curve parameters
##### MARKER - add figure of this
We estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).

$$\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx$$
where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively. We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of \aax{} for each leaf at each light intensity. We then summarized the posterior distribution of \aax{} for each leaf at each light intensity by the posterior median, which became our estimate of \aax{} for subsequent analyses.

### Bayesian phylogenetic mixed effects models

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We crossed those fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagnse*; LA0750 was added as sister to LA0716, two closely replated populations of *S. pennellii*. The node was placed half-way between the next deepest node.

NEED TO EDIT THIS: The next sections describe how we fit models in *Stan*, all model parameters and priors, and specific predictions about parameter values for each hypothesis.
Each curve was fit For initial fits, we sampled the posterior distribution from a single chain with `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. After model selection (see below), we refit the selected model with more iterations if necessary to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.

#### Fitting models in *Stan* using **brms** interface {#sec-fitting}

We fit Bayesian models with HMC sampling in the probabilistic programming language *Stan* [@stan_development_team_stan_2025] using the *R* package **brms** version `r fit_aa1$version$brms` [@burkner_brms_2017]. We used *CmdStan* version `r fit_aa1$version$cmdstan` and **cmdstanr** version `r fit_aa1$version$cmdstanr` [@gabry_cmdstanr_2025] to interface with *R* version `r getRversion()` [@r_core_team_r:_2025]. For initial fits, we sampled the posterior distribution from a single chain with `r aa_args$n_iter_init / 2` iterations after `r aa_args$n_iter_init / 2` warmup iterations. After model selection (see below), we refit the selected model with more iterations if necessary to decrease the Gelman-Rubin convergence statistic ($\hat{R}$) [@gelman_inference_1992] to less than $`aa_stats[["max_rhat"]]`$ and the bulk effective sample size (ESS) to greater than $`aa_stats[["min_ess"]]`$ for all model parameters.

##### Effects of light intensity, light treatment, and population on \aax

We tested for effects of light intensity, light treatment, population, and their interactions on \aax. All models included effects of light intensity and light treatment, as well as random effects of population and replicate within population. More complex models included interactions between light intensity and light treatment, as well as random effects of population on the effects of light intensity and light treatment. We used a weakly informative normal prior with mean $0$ and standard deviation $10$ for fixed effects of light intensity and treatment. We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ for the the random effect standard deviation of replicate within population. We accounted for the phylogenetic structure among the random effects of population using an Ornstein-Uhlenbeck (OU) process [@hansen_stabilizing_1997]. The expected covariance between populations $i$ and $j$ ($\text{Cov}(i, j)$ ) is:

$$\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})$$
where $\sigma^2$ is the variance of the random effect, $\alpha$ is the rate of decay of the covariance with phylogenetic distance, $D_{ij}$, the phylogenetic distance between populations $i$ and $j$. We estimated $\sigma^2 / (2 \alpha)$ and $\alpha$ as a separate parameters and reparameterized them as $\sigma^2$ and $\alpha$. We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.

We modeled the residual standard deviation of \aax, (i.e. phylogenetically unstructured variation unaccounted for by explanatory variables) on a log-link scale with effects of light intensity and light treatment. We used a weakly informative normal prior with mean $-3$ and standard deviation $5$ on the residual standard deviation intercept and a weakly informative normal prior with mean $0$ and standard deviation $1$ on the effects of light intensity and light treatment on the residual standard deviation.

##### Effects of native light habitat on population-level \aax

We tested a linear effect of native light habitat on population-level \aax. In Models 1 and 2, we used the random intercept of \aax{} as a response variable; in model 3 we used the random intercept plus the random effect of population at high light intensity; in model 4 we used the random intercept plus the random effect of population in the sun treatment; in model 5 we used the random intercept plus the random effects of population at high light intensity and sun treatment. We used a weakly informative normal prior with mean $0$ and standard deviation $1$ for the slope and intercept. We accounted for the phylogenetic structure among the model residuals using an (OU) process. We used a weakly informative normal prior with mean $0$ and standard deviation $10$ on OU parameters.

#### Predictions {#sec-predictions}

@@

We evaluated predictions using a combination of parameter estimation and model selection. In all cases, we estimated parameters and confidence intervals as described in @sec-fitting. If the 95% confidence intervals for a parameter did not overlap zero, we considered the parameter to be significantly different from zero. We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values. Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{} (see @sec-parameters). We considered models with two standard errors of the mean lower LOOIC value to be a better fit to the data; we considered models with LOOIC values within two standard errors of the mean to be have similar support.

@@

>>>>

#### \agcurve{} curve parameters {#sec-agcurve-pars}

@@

#### \aax{} for each light intensity within leaf using \agcurve{} curve parameters {#sec-aa-est}

We estimated \aax{} for each light intensity by integrating the difference in \logA{} between the amphi and pseudohypo \agcurve{} curves over the range of \gsw{} values where the curves overlap (from $\min(\log(g_\text{sw}))$ to $\max(\log(g_\text{sw}))$).

$$\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx$$
where $\theta_\text{amphi}$ and $\theta_\text{hypo}$ are the quadratic parameters of the amphi and pseudohypo \agcurve{} curves, respectively (@fig-example-aa). We calculated $\widehat{\mathrm{AA}}$ from each draw of the posterior for the paired amphi and pseudohypo \agcurve{} curves to obtain the posterior distribution of \aax{} for each leaf at each light intensity. We then summarized the posterior distribution of \aax{} for each leaf at each light intensity by the posterior median and standard deviation, which became our point estimate of, and uncertainty in, \aax{} for the phylogenetic mixed effects model described below.

```{r, example-aa, echo=FALSE,message=FALSE,warning=FALSE}
#| fig-align: center
#| fig-width: 6.5
#| fig-height: 3.5
#| fig-cap: "Example amphistomy advantage (\\aax) calculation from \\agcurve{} curves in an amphistomatous leaf with gas exchange through both surfaces (blue) and the same leaf with gas exchange blocked on the upper surface only (orange). We integrate over the gray region between the curves to estimate \\aax{}. The \\aax{} is low (left facet) when the photosynthetic rate of an amphistomatous leaf is similar to a pseudohypostomatous leaf at the same total stomatal conductance ($x$-axes); large \\aax{} (right facet) is indicated when an amphistomatous leaf has a higher photosynthetic rate than a pseudohypostomatous leaf."
read_rds("../figures/example-aa.rds")
```

#### Phylogenetic mixed effects models {#sec-phylo}

We fit Bayesian mixed effects models with phylogenetically structured random effects to test predictions of competing hypotheses about why amphistomy advantage (\aax) might be greater for leaves in sunny, open habitats. We fit all models combinations of measurement light intensity, growth light intensity, and their interaction as both fixed effects and random effects among populations. In other words, we test whether there were main effects of light intensity treatments and/or whether populations varied in their response to light intensity treatments. In all models, we included population as a phylogenetically structured random effect. We accounted for uncertainty in estimates of \aax{} using the standard deviation of the posterior distribution of each \aax{} estimate (@sec-aa-est). We crossed fixed and random effect structures with fixed effects of measurement light intensity and/or growth light intensity on the distributional parameter, $\sigma$, the residual variance in \aax{}. In total, we fit `r n_model1` models, each with a different combination of fixed, random, and distributional effects. We used the RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb genomic windows from @pease_phylogenomics_2016. Two of our populations were not in this tree. We used accession LA1044 in place of LA3909, two populations of *S. galapagnse*; LA0750 was added as sister to LA0716, two closely related populations of *S. pennellii*. The node separating LA0750 from LA0716 was placed half-way between the next deepest node. We used the leave-one-out cross-validation information criterion (LOOIC) to compare the fit of models (\autoref{tbl-models}) using the *R* package **loo** version `r packageVersion("loo")` [@vehtari_practical_2017] to calculate LOOIC values. We selected the model with the lowest LOOIC value to generate posterior predictions for hypothesis testing.

#### Posterior predictions {#sec-predictions}

@@

From the posterior of the selected model (@sec-phylo), we predicted the posterior distribution of the expected \aax{} for each population at each measurement light intensity and growth light intensity. If the 95% confidence intervals for \aax{} did not overlap zero, we considered \aax{} to be significantly different from zero. We also calculated posterior distribution for the average \aax{} among populations in each condition. We tested a linear effect of native light habitat on population-level \aax{} in each treatment condition, integrating over the entire posterior distribution. We calculated the posterior distribution of the slope. If the 95\% confidence interval for the slope did not overlap zero, we considered the slope to be significant.

## MARKER - need to update table

@@

THIS COULD BE USED FOR LMA MODEL? Note that LOOIC was calculated only from pointwise likelihood values of the submodel estimating effects of light intensity, light treatments, and population on \aax{}.

### 2025-07-20 ms/ms.tex
example-aa figure

\pandocbounded{\includegraphics[keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}}

@@

Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). Bayesian
phylogenetic mixed effects models that allowed \aax{} to vary between
measurement light intensities, growth light intensities, and among
populations outperformed simpler models based on their information
criteria (Table~\ref{tbl-aa_loo1}). Measured under high light intensity,
\aax{} was consistently greater for sun plants. The average \aax{} among
populations in the shade treatment was 0.041 (range: 0.010--0.113; 21 of
29 populations significant); however, the same populations grown at high
light intensity showed a mean \aax{} of 0.053 (range: 0.021--0.125; 19
of 29 populations significant). Contrary to the predictions of the
assimilatory hypothesis, \aax{} was greater in all populations under low
measurement light intensity for both sun and shade grown plants. The
effect of low light on \aax{} was more pronounced in the sun-grown
plants, where \aax{} was significantly greater under low measurement
light intensity in 11 of 29 populations compared to to 4 populations for
shade-grown plants. The overall average \aax{} of shade and sun grown
plants measured under low light intensity was 0.064 (range:
0.017--0.132; 28 of 29 populations significant) and 0.094 (range:
0.036--0.197; 24 of 29 populations significant), respectively. There was
a slight tendency for populations from more closed habitats (greater
PAI), but \aax{} varied widely among populations from open habitats (low
PAI) regardless of growth and measurement light intensities
(Fig.~\ref{fig-aa}B).

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, argues against the acclimatory hypothesis, and
provides only weak support for the constitutive hypothesis. Plastic
changes in leaf thickness and/or packing density, summarized by the bulk
leaf mass per area (LMA), likely mediate the effect of growth light
intensity on \aax{}. LMA increased in sun grown plants in all
populations by an average of 123\% {[}95\% CI: 50.5 to 245\%{]}, similar
to plastic responses in many species (\emph{48}). When LMA is included
in the model, the effect of growth light intensity on \aax{} is no
longer significant based on model comparison (Table S8) and LMA is
weakly positively correlated with \aax{} (Fig.~\ref{fig-lma_aa}). In
contrast, LMA did not mediate an indirect effect of native PAI on \aax{}
(Table S8). Many anatomical traits underlie LMA (\emph{53}) and future
research will be needed identify which particular traits, such as leaf
thickness or mesophyll porosity, are responsible for mediating \aax{}.
The fact that the \aax{} of sun plants was greater under low measurement
light intensity supports a long-standing hypothesis that resistance to
CO\(_2\) diffusion is greater in the upper than lower portions of the
leaf interior (\emph{54}). At low light intensity, the bulk properties
of the leaf are weighed toward the upper palisade where most light is
intercepted. Hence, amphistomy may, expectedly, be particularly
beneficial for sun leaves experiencing intermittant shade or cloud
cover. Our study is limited in testing this because we could not
directly measure the stomatal conductance ratio and intercellular
resistance on each surface. Future experiments measuring \aax{} with a
dual sided chamber (\emph{31}, \emph{33}) can overcome these
limitations.

@@

\bibitem[\citeproctext]{ref-hansen_stabilizing_1997}
\CSLLeftMargin{64. }%
\CSLRightInline{T. F. Hansen,
\href{https://doi.org/10.1111/j.1558-5646.1997.tb01457.x}{Stabilizing
selection and the comparative analysis of adaptation}. \emph{Evolution}
\textbf{51}, 1341--1351 (1997).}

@@

\paragraph{\texorpdfstring{\agcurve{} curve
parameters}{ curve parameters}}\label{curve-parameters}

@@

\paragraph{\texorpdfstring{\aax{} for each light intensity within leaf
using \agcurve{} curve
parameters}{ for each light intensity within leaf using  curve parameters}}\label{for-each-light-intensity-within-leaf-using-curve-parameters}

\subparagraph{MARKER - add figure of
this}\label{marker---add-figure-of-this}

@@

\[\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx\]
where \(\theta_\text{amphi}\) and \(\theta_\text{hypo}\) are the
quadratic parameters of the amphi and pseudohypo \agcurve{} curves,
respectively. We calculated \(\widehat{\mathrm{AA}}\) from each draw of
the posterior for the paired amphi and pseudohypo \agcurve{} curves to
obtain the posterior distribution of \aax{} for each leaf at each light
intensity. We then summarized the posterior distribution of \aax{} for
each leaf at each light intensity by the posterior median, which became
our estimate of \aax{} for subsequent analyses.

\subsubsection{Bayesian phylogenetic mixed effects
models}\label{bayesian-phylogenetic-mixed-effects-models}

We fit Bayesian mixed effects models with phylogenetically structured
random effects to test predictions of competing hypotheses about why
amphistomy advantage (\aax) might be greater for leaves in sunny, open
habitats. We fit all models combinations of measurement light intensity,
growth light intensity, and their interaction as both fixed effects and
random effects among populations. In other words, we test whether there
were main effects of light intensity treatments and/or whether
populations varied in their response to light intensity treatments. In
all models, we included population as a phylogenetically structured
random effect. We crossed those fixed and random effect structures with
fixed effects of measurement light intensity and/or growth light
intensity on the distributional parameter, \(\sigma\), the residual
variance in \aax{}. In total, we fit 100 models, each with a different
combination of fixed, random, and distributional effects. We used the
RAxML whole-transcriptome concatenated phylogeny based on 2,745 100-kb
genomic windows from (\textbf{pease\_phylogenomics\_2016?}). Two of our
populations were not in this tree. We used accession LA1044 in place of
LA3909, two populations of \emph{S. galapagnse}; LA0750 was added as
sister to LA0716, two closely replated populations of \emph{S.
pennellii}. The node was placed half-way between the next deepest node.

NEED TO EDIT THIS: The next sections describe how we fit models in
\emph{Stan}, all model parameters and priors, and specific predictions
about parameter values for each hypothesis. Each curve was fit For
initial fits, we sampled the posterior distribution from a single chain
with 1000 iterations after 1000 warmup iterations. After model selection
(see below), we refit the selected model with more iterations if
necessary to decrease the Gelman-Rubin convergence statistic
(\(\hat{R}\)) (\emph{63}) to less than \(`aa_stats[["max_rhat"]]`\) and
the bulk effective sample size (ESS) to greater than
\(`aa_stats[["min_ess"]]`\) for all model parameters.

\paragraph{\texorpdfstring{Fitting models in \emph{Stan} using
\textbf{brms}
interface}{Fitting models in Stan using brms interface}}\label{sec-fitting}

We fit Bayesian models with HMC sampling in the probabilistic
programming language \emph{Stan} (\emph{59}) using the \emph{R} package
\textbf{brms} version 2.22.0 (\emph{60}). We used \emph{CmdStan} version
2.36.0 and \textbf{cmdstanr} version 0.9.0 (\emph{61}) to interface with
\emph{R} version 4.5.0 (\emph{62}). For initial fits, we sampled the
posterior distribution from a single chain with 1000 iterations after
1000 warmup iterations. After model selection (see below), we refit the
selected model with more iterations if necessary to decrease the
Gelman-Rubin convergence statistic (\(\hat{R}\)) (\emph{63}) to less
than \(`aa_stats[["max_rhat"]]`\) and the bulk effective sample size
(ESS) to greater than \(`aa_stats[["min_ess"]]`\) for all model
parameters.

\subparagraph{\texorpdfstring{Effects of light intensity, light
treatment, and population on
\aax}{Effects of light intensity, light treatment, and population on }}\label{effects-of-light-intensity-light-treatment-and-population-on}

We tested for effects of light intensity, light treatment, population,
and their interactions on \aax. All models included effects of light
intensity and light treatment, as well as random effects of population
and replicate within population. More complex models included
interactions between light intensity and light treatment, as well as
random effects of population on the effects of light intensity and light
treatment. We used a weakly informative normal prior with mean \(0\) and
standard deviation \(10\) for fixed effects of light intensity and
treatment. We used a weakly informative normal prior with mean \(-3\)
and standard deviation \(5\) for the the random effect standard
deviation of replicate within population. We accounted for the
phylogenetic structure among the random effects of population using an
Ornstein-Uhlenbeck (OU) process (\emph{64}). The expected covariance
between populations \(i\) and \(j\) (\(\text{Cov}(i, j)\) ) is:

\[\text{Cov}(i, j) = \frac{\sigma^2}{2 \alpha} \exp(-\alpha D_{ij})\]
where \(\sigma^2\) is the variance of the random effect, \(\alpha\) is
the rate of decay of the covariance with phylogenetic distance,
\(D_{ij}\), the phylogenetic distance between populations \(i\) and
\(j\). We estimated \(\sigma^2 / (2 \alpha)\) and \(\alpha\) as a
separate parameters and reparameterized them as \(\sigma^2\) and
\(\alpha\). We used a weakly informative normal prior with mean \(0\)
and standard deviation \(10\) on OU parameters.

We modeled the residual standard deviation of \aax,
(i.e.~phylogenetically unstructured variation unaccounted for by
explanatory variables) on a log-link scale with effects of light
intensity and light treatment. We used a weakly informative normal prior
with mean \(-3\) and standard deviation \(5\) on the residual standard
deviation intercept and a weakly informative normal prior with mean
\(0\) and standard deviation \(1\) on the effects of light intensity and
light treatment on the residual standard deviation.

\subparagraph{\texorpdfstring{Effects of native light habitat on
population-level
\aax}{Effects of native light habitat on population-level }}\label{effects-of-native-light-habitat-on-population-level}

We tested a linear effect of native light habitat on population-level
\aax. In Models 1 and 2, we used the random intercept of \aax{} as a
response variable; in model 3 we used the random intercept plus the
random effect of population at high light intensity; in model 4 we used
the random intercept plus the random effect of population in the sun
treatment; in model 5 we used the random intercept plus the random
effects of population at high light intensity and sun treatment. We used
a weakly informative normal prior with mean \(0\) and standard deviation
\(1\) for the slope and intercept. We accounted for the phylogenetic
structure among the model residuals using an (OU) process. We used a
weakly informative normal prior with mean \(0\) and standard deviation
\(10\) on OU parameters.

\paragraph{Predictions}\label{sec-predictions}

@@

We evaluated predictions using a combination of parameter estimation and
model selection. In all cases, we estimated parameters and confidence
intervals as described in Section~\ref{sec-fitting}. If the 95\%
confidence intervals for a parameter did not overlap zero, we considered
the parameter to be significantly different from zero. We used the
leave-one-out cross-validation information criterion (LOOIC) to compare
the fit of models (\autoref{tbl-models}) using the \emph{R} package
\textbf{loo} version 2.8.0 (\emph{65}) to calculate LOOIC values. Note
that LOOIC was calculated only from pointwise likelihood values of the
submodel estimating effects of light intensity, light treatments, and
population on \aax{} (see \textbf{?@sec-parameters}). We considered
models with two standard errors of the mean lower LOOIC value to be a
better fit to the data; we considered models with LOOIC values within
two standard errors of the mean to be have similar support.

>>>>

\includegraphics[width=2.24in,height=\textheight,keepaspectratio]{../figures/amphisomy draft 2024-03-28.jpg}

@@

Consistent with biophysical theory of CO\(_2\) diffusion within leaves,
\(\aay{} > 0\) for all populations (Fig.~\ref{fig-aa}A). Bayesian
phylogenetic mixed effects models that allowed \aax{} to vary between
measurement light intensities, growth light intensities, and among
populations outperformed simpler models based on their information
criteria (Table~\ref{tbl-aa_loo1}). Measured under high light intensity,
\aax{} was consistently greater for sun plants. The average \aax{} among
populations in the shade treatment was 0.041 (range: 0.010--0.113; 21 of
29 populations significant); however, the same populations grown at high
light intensity showed a mean \aax{} of 0.053 (range: 0.021--0.125; 19
of 29 populations significant). Contrary to the predictions of the
assimilatory hypothesis, \aax{} was greater in all populations under low
measurement light intensity for both sun and shade grown plants. The
effect of low light on \aax{} was more pronounced in the sun-grown
plants, where \aax{} was significantly greater under low measurement
light intensity in 11 of 29 populations compared to 4 populations for
shade-grown plants. The overall average \aax{} of shade and sun grown
plants measured under low light intensity was 0.064 (range:
0.017--0.132; 28 of 29 populations significant) and 0.094 (range:
0.036--0.197; 24 of 29 populations significant), respectively. There was
a slight tendency for populations from more closed habitats (greater
PAI), but \aax{} varied widely among populations from open habitats (low
PAI) regardless of growth and measurement light intensities
(Fig.~\ref{fig-aa}B).

@@

The pattern of \aax{} across wild tomatoes strongly supports the
plasticity hypothesis, argues against the acclimatory hypothesis, and
provides only weak support for the constitutive hypothesis. Plastic
changes in leaf thickness and/or packing density, summarized by the bulk
leaf mass per area (LMA), likely mediate the effect of growth light
intensity on \aax{}. LMA increased in sun grown plants in all
populations by an average of 119\% {[}95\% CI: 47 to 240\%{]}, similar
to plastic responses in many species (\emph{48}). When LMA is included
in the model, the effect of growth light intensity on \aax{} is no
longer significant based on model comparison (Table S8) and LMA is
weakly positively correlated with \aax{} (Fig.~\ref{fig-lma_aa}). In
contrast, LMA did not mediate an indirect effect of native PAI on \aax{}
(Table S8). Many anatomical traits underlie LMA (\emph{53}) and future
research will be needed identify which particular traits, such as leaf
thickness or mesophyll porosity, are responsible for mediating \aax{}.
The fact that the \aax{} of sun plants was greater under low measurement
light intensity supports a long-standing hypothesis that resistance to
CO\(_2\) diffusion is greater in the upper than lower portions of the
leaf interior (\emph{54}). At low light intensity, the bulk properties
of the leaf are weighed toward the upper palisade where most light is
intercepted. Hence, amphistomy may, expectedly, be particularly
beneficial for sun leaves experiencing intermittant shade or cloud
cover. Our study is limited in testing this because we could not
directly measure the stomatal conductance ratio and intercellular
resistance on each surface. Future experiments measuring \aax{} with a
dual sided chamber (\emph{31}, \emph{33}) can overcome these
limitations.

@@

\bibitem[\citeproctext]{ref-pease_phylogenomics_2016}
\CSLLeftMargin{64. }%
\CSLRightInline{J. B. Pease, D. C. Haak, M. W. Hahn, L. C. Moyle,
\href{https://doi.org/10.1371/journal.pbio.1002379}{Phylogenomics
reveals three sources of adaptive variation during a rapid radiation}.
\emph{PLOS Biology} \textbf{14}, e1002379 (2016).}

@@

\paragraph{\texorpdfstring{\agcurve{} curve
parameters}{ curve parameters}}\label{sec-agcurve-pars}

@@

\paragraph{\texorpdfstring{\aax{} for each light intensity within leaf
using \agcurve{} curve
parameters}{ for each light intensity within leaf using  curve parameters}}\label{sec-aa-est}

@@

\[\widehat{\mathrm{AA}} = \int_{\text{min(log(}g_\text{sw}))}^{\text{max(log(}g_\text{sw}))} \text{log}\bigg(\frac{\hat{A}_\text{amphi}(x; \theta_{\text{amphi}})}{\hat{A}_\text{hypo}(x; \theta_{\text{hypo}})}\bigg) dx\]
where \(\theta_\text{amphi}\) and \(\theta_\text{hypo}\) are the
quadratic parameters of the amphi and pseudohypo \agcurve{} curves,
respectively (\textbf{?@fig-example-aa}). We calculated
\(\widehat{\mathrm{AA}}\) from each draw of the posterior for the paired
amphi and pseudohypo \agcurve{} curves to obtain the posterior
distribution of \aax{} for each leaf at each light intensity. We then
summarized the posterior distribution of \aax{} for each leaf at each
light intensity by the posterior median and standard deviation, which
became our point estimate of, and uncertainty in, \aax{} for the
phylogenetic mixed effects model described below.

\begin{figure}[H]

{\centering \pandocbounded{\includegraphics[keepaspectratio]{ms_files/figure-pdf/example-aa-1.pdf}}

\caption{Example amphistomy advantage (\aax) calculation from \agcurve{}
curves in an amphistomatous leaf with gas exchange through both surfaces
(blue) and the same leaf with gas exchange blocked on the upper surface
only (orange). We integrate over the gray region between the curves to
estimate \aax{}. The \aax{} is low (left facet) when the photosynthetic
rate of an amphistomatous leaf is similar to a pseudohypostomatous leaf
at the same total stomatal conductance (\(x\)-axes); large \aax{} (right
facet) is indicated when an amphistomatous leaf has a higher
photosynthetic rate than a pseudohypostomatous leaf.}

\paragraph{Phylogenetic mixed effects models}\label{sec-phylo}

We fit Bayesian mixed effects models with phylogenetically structured
random effects to test predictions of competing hypotheses about why
amphistomy advantage (\aax) might be greater for leaves in sunny, open
habitats. We fit all models combinations of measurement light intensity,
growth light intensity, and their interaction as both fixed effects and
random effects among populations. In other words, we test whether there
were main effects of light intensity treatments and/or whether
populations varied in their response to light intensity treatments. In
all models, we included population as a phylogenetically structured
random effect. We accounted for uncertainty in estimates of \aax{} using
the standard deviation of the posterior distribution of each \aax{}
estimate (Section~\ref{sec-aa-est}). We crossed fixed and random effect
structures with fixed effects of measurement light intensity and/or
growth light intensity on the distributional parameter, \(\sigma\), the
residual variance in \aax{}. In total, we fit 100 models, each with a
different combination of fixed, random, and distributional effects. We
used the RAxML whole-transcriptome concatenated phylogeny based on 2,745
100-kb genomic windows from (\emph{64}). Two of our populations were not
in this tree. We used accession LA1044 in place of LA3909, two
populations of \emph{S. galapagnse}; LA0750 was added as sister to
LA0716, two closely related populations of \emph{S. pennellii}. The node
separating LA0750 from LA0716 was placed half-way between the next
deepest node. We used the leave-one-out cross-validation information
criterion (LOOIC) to compare the fit of models (\autoref{tbl-models})
using the \emph{R} package \textbf{loo} version 2.8.0 (\emph{65}) to
calculate LOOIC values. We selected the model with the lowest LOOIC
value to generate posterior predictions for hypothesis testing.

\paragraph{Posterior predictions}\label{sec-predictions}

@@

From the posterior of the selected model (Section~\ref{sec-phylo}), we
predicted the posterior distribution of the expected \aax{} for each
population at each measurement light intensity and growth light
intensity. If the 95\% confidence intervals for \aax{} did not overlap
zero, we considered \aax{} to be significantly different from zero. We
also calculated posterior distribution for the average \aax{} among
populations in each condition. We tested a linear effect of native light
habitat on population-level \aax{} in each treatment condition,
integrating over the entire posterior distribution. We calculated the
posterior distribution of the slope. If the 95\% confidence interval for
the slope did not overlap zero, we considered the slope to be
significant.

\subsection{MARKER - need to update
table}\label{marker---need-to-update-table}
